---
title: "Ferrari Test"
author: "July Pilowsky"
format: pdf
---

```{r setup, echo = F, include = F}
library(tidyverse)
library(scales)
library(paletteer)
library(cowplot)
theme_set(theme_cowplot())
set.seed(185)
```

# What This Test Does

The scenario is that there are two viruses, A and B, that cause permanent infection. Infection is through frequency-dependent direct transmission. There are two scenarios: *facilitation*, in which the strains each facilitate infection with the other, and *competition*, in which each strain suppresses infection with the other. The facilitation and competition are both symmetrical.

The host population is 1000 individuals. I will start with two infected individuals: one with strain A, and one with strain B. I could use differential equations for this but I will do it discretely with demographic stochasticity (though there will be no demography in this population.)

# Setup

For both scenarios, the starting host population is the same. I will represent the host population as a named vector representing 4 population compartments: susceptible individuals, S; those infected with strain A only, I~a~; those infected with strain B only, I~b~; and those infected with both, I~ab~.

```{r initial population}
init <- c(S = 998, Ia = 1, Ib = 1, Iab = 0)
```

I will create a set of parameters with 𝛃~S~ (transmission of both strains to susceptibles) and 𝛃~I~ (transmission of each strain to those infected with the other.) For the *competition* scenario, 𝛃~S~ \> 𝛃~I~, and for the *facilitation* scenario, 𝛃~S~ \< 𝛃~I~.

```{r scenario parms}
competition_params <- c(beta_s = 0.0003, beta_i = 0.0002)
facilitation_params <- c(beta_s = 0.0003, beta_i = 0.0004)
```

# Simulation

Here is the function I will use for simulations.

```{r simulation function}
ferrari <- function(initial_vector, params, timesteps) {
  # Unpack parameters
  beta_s <- params["beta_s"]
  beta_i <- params["beta_i"]
  
  # Set up population vector
  state_list <- c(list(initial_vector), replicate(timesteps, c(
    S = NA,
    Ia = NA,
    Ib = NA,
    Iab = NA
  ), simplify = F))
  
  for (t in 1:timesteps) {
    
    # Unpack states
    S <- state_list[[t]]["S"]
    Ia <- state_list[[t]]["Ia"]
    Ib <- state_list[[t]]["Ib"]
    Iab <- state_list[[t]]["Iab"]
    
    if (Iab == 1000) {
      state_list[[t + 1]] <- state_list[[t]]
    } else {
      # State changes
      infection_s_a <- rbinom(1, S*(Ia + Iab), beta_s)
      infection_s_b <- rbinom(1, S*(Ib + Iab), beta_s)
      infection_a_ab <- rbinom(1, Ia*(Ib + Iab), beta_i)
      infection_b_ab <- rbinom(1, Ib*(Ia + Iab), beta_i)
      infection_s <- infection_s_a + infection_s_b
      if (infection_s > S) {
        ratio <- infection_s_a/(infection_s_a + infection_s_b)
        infection_s_a <- round(S*ratio)
        infection_s_b <- round(S*(1-ratio))
      }
      if (infection_b_ab > Ib) {
        infection_b_ab <- Ib
      }
      if (infection_a_ab > Ia) {
        infection_a_ab <- Ia
      }
      
      # Store new states
      state_list[[t + 1]]["S"] <- S - infection_s_a - infection_s_b
      state_list[[t + 1]]["Ia"] <- Ia + infection_s_a - infection_a_ab
      state_list[[t + 1]]["Ib"] <- Ib + infection_s_b - infection_b_ab
      state_list[[t + 1]]["Iab"] <- Iab + infection_a_ab + infection_b_ab
    }
  }
  
  df <- state_list %>% bind_rows() %>% rowid_to_column("Timestep")
  return(df)
}
```

Here I use it for the *competition* scenario.

```{r competition scenario}
competition_df <- ferrari(init, competition_params, 100)
competition_df %>% 
  pivot_longer(S:Iab, names_to = "State", values_to = "Count") %>% 
  ggplot(aes(x = Timestep, y = Count, group = State)) +
  geom_line(aes(col = State), linewidth = 1.05) +
  scale_color_paletteer_d(`"fishualize::Coryphaena_hippurus"`) +
  theme_minimal() +
  ggtitle("Competition")
```

And here is the *facilitation* scenario.

```{r facilitation scenario}
facilitation_df <- ferrari(init, facilitation_params, 100)
facilitation_df %>% 
  pivot_longer(S:Iab, names_to = "State", values_to = "Count") %>% 
  ggplot(aes(x = Timestep, y = Count, group = State)) +
  geom_line(aes(col = State), linewidth = 1.05) +
  scale_color_paletteer_d(`"fishualize::Coryphaena_hippurus"`) +
  theme_minimal() +
  ggtitle("Facilitation")
```

Once again, I can summarize the differences between the scenarios by comparing how many timesteps it takes to get to half the population infected with both strains:

```{r summary metric}
competition_df %>% pull(Iab) %>% map_lgl(~.>500) %>% which.max() %>% 
  set_names("competition")
facilitation_df %>% pull(Iab) %>% map_lgl(~.>500) %>% which.max() %>% 
  set_names("facilitation")
```

# Sampling

We are interested in how these interactive effects change what you observe in a disease system when you sample. So let's look at each scenario sampled at timestep 20 and timestep 40. This is a case of perfect sampling, where the whole population is surveyed at each timepoint and the test used is totally accurate.

```{r perfect sampling}
competition_df[c(20,40),] %>% mutate(Scenario = "Competition") %>% 
  bind_rows(facilitation_df[c(20,40),] %>% mutate(Scenario = "Facilitation")) %>% 
  pivot_longer(S:Iab, names_to = "Infection_Status", values_to = "Count") %>% 
  ggplot(aes(x = Scenario, y = Count, fill = Infection_Status)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~Timestep) +
  scale_fill_paletteer_d("fishualize::Coryphaena_hippurus", 
                         name = "Infection\nStatus")
```

# The Multi-Species Case

Now I will adapt this model so it can take any number of species. To do this, I will first need to generate a matrix of interactions among strains: multipliers that modify transmission parameters.

```{r interaction matrix}
beta_s <- 0.0003
interactions <- runif(25, 0.5, 1.5) |> matrix(nrow = 5)
diag(interactions) <- 1
interactions
```

We also need to store the initial population in a matrix that can keep track of which strains the population has or doesn't have. I include one individual infected with each of the five strains.

```{r initial population 2}
initial_pop <- c(rep(c(1, rep(0, 5)), 5), rep(0, 4970)) |> 
  matrix(nrow = 1000, byrow = T)
```

Here is a simulation function that can handle any number of strains with an interaction matrix, assuming their baseline transmission rate is the same:

```{r simulation function 2}
ferrari <- function(initial_pop,
                    interactions,
                    beta_s,
                    timesteps) {
  # Set up population list
  state_list <-
    c(list(), replicate(timesteps, matrix(
      0,
      ncol = ncol(initial_pop),
      nrow = nrow(initial_pop)
    ), simplify = FALSE))
  state_list[[1]] <- initial_pop
  
  for (t in 1:(timesteps - 1)) {
    current_pop <- state_list[[t]]
    col_sums <- colSums(current_pop)
    
    if (sum(col_sums) == length(current_pop)) {
      state_list[[t + 1]] <- current_pop
    } else {
      indiv <- array_branch(current_pop, 1)
      
      total_strains <- map_dbl(indiv, ~ sum(.x))
      
      indiv_new <- map(1:length(indiv), \(i) {
        indiv_prev <- indiv[[i]]
        int_indiv <- interactions[as.logical(indiv_prev),]
        
        indiv_new <- map_dbl(1:length(indiv_prev), \(column) {
          if (total_strains[i] > 1) {
            pmax(indiv_prev[column],
                 rbinom(
                   1,
                   col_sums[column],
                   int_indiv[, column] |>
                     Reduce(`*`, x = _) |> (`*`)(beta_s)
                 ))
          } else if (total_strains[i] == 1) {
            pmax(indiv_prev[column],
                 rbinom(1,
                        col_sums[column],
                        int_indiv[column] |> (`*`)(beta_s)))
          } else {
            rbinom(1, col_sums[column], beta_s)
          }
        })
        
        matrix(pmax(indiv_prev, indiv_new |> map_dbl(\(x) {
          ifelse(x > 1, 1, x)
        })), nrow = 1)
      })
      
      state_list[[t + 1]] <-
        matrix(unlist(indiv_new),
               nrow = nrow(current_pop),
               byrow = T)
    }
  }
  
  
  df <- state_list |> map(colSums) |>
    map(\(x) {
      set_names(x, LETTERS[1:ncol(initial_pop)])
    }) |>
    bind_rows() |>
    rowid_to_column("Timestep")
  return(df)
}
```

Now I will try out my randomly generated interaction matrix.

```{r five strain test}
five_strain <- ferrari(initial_pop, interactions, beta_s, 100)
five_strain |> pivot_longer(A:E, names_to = "Strain", values_to = "Cases") |> 
  ggplot(aes(x = Timestep, y = Cases, group = Strain)) +
  geom_line(aes(col = Strain), linewidth = 1.05, alpha = 0.7) +
  scale_color_paletteer_d("dichromat::Categorical_12", -1)
```
